---
title: "Project 2"
author: "Thomas Gresham, Eric Jess, Harrison Mazanec"
date: "12/7/2020"
pledge: "On our honor, we have neither given nor recieved aid on this assignment."
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
tinytex::install_tinytex()
require("knitr")
sourcedir <- "C:/Users/etj3pp/Desktop/SYS4021/Source"
datadir <- "C:/Users/etj3pp/Desktop/SYS4021/Data/AirQualityUCI"
opts_knit$set(root.dir = sourcedir)

# load libraries
library(ggplot2)
library(ggfortify)
library(ggpubr)

library(forecast)
library(imputeTS)
library(lubridate)
library(tidyverse)
library(mtsdi)
library(tseries)

library("car")
```

Load data and impute missing values
```{r load data}
setwd(datadir)

# Read in data file
airquality = read.csv('AirQualityUCI.csv')

# Replace -200 with NA
airquality[airquality == -200] <- NA

# Convert integer type to numeric
intcols = c(4,5,7,8,9,10,11,12)
for(i in 1:length(intcols)){
  airquality[,intcols[i]] <- as.numeric(airquality[,intcols[i]])
}

setwd(sourcedir)

# Create new data frame with just NO2 and impute missing values
AQdata = airquality["NO2.GT."]
AQdata = na_interpolation(AQdata)

# Aggregate to daily maxima for model building
dailyAQ <- aggregate(AQdata, by=list(as.Date(airquality[,1],"%m/%d/%Y")), FUN=max)

# Create time series of NO2
orig.NO2.ts <- ts(dailyAQ[,2])

# Remove last 7 days of observations
dailyAQ.new <- dailyAQ[1:(dim(dailyAQ)[1]-7),]
NO2.ts <- ts(dailyAQ.new[,2])

```

Part 1A - Seasonal Components
```{r 1A}

# Plot time series data
autoplot(NO2.ts, ylab = "Maximum Daily Nitrogen Dioxide Concentration", xlab = "Day")

# Periodogram of time series data
pg.NO2 <- spec.pgram(log(NO2.ts), spans = 9, demean=T, log='no')

# Peak of periodogram
max.omega.NO2 <- pg.NO2$freq[which(pg.NO2$spec==max(pg.NO2$spec))]
max.omega.NO2 #0.00520833

# Period of data 
1/max.omega.NO2 #192 days

# Sort spectrum from largest to smallest and find index
sorted.spec <- sort(pg.NO2$spec, decreasing=T, index.return=T)
names(sorted.spec)

# Corresponding periods (omegas = frequences, Ts = periods)
sorted.omegas <- pg.NO2$freq[sorted.spec$ix]
sorted.Ts <- 1/pg.NO2$freq[sorted.spec$ix]

# Look at first 20 frequencies
sorted.omegas[1:20]
sorted.Ts[1:20]

# From looking at the periodogram and maximum peaks, there is no clear period. However,
# NO2 may depends on day of the week (travel, work commute, etc.)

# Model seasonality based on days of the week (like we did for ham in TS1)
t <- c(seq(1:dim(dailyAQ.new)[1]))

# March 10, 2004 is a Wednesday (first day in data frame)
Day <- rep(NA, length(NO2.ts))
Day[which((t %% 7)    == 1)] <- "W" #wednesday
Day[which((t %% 7)    == 2)] <- "R" #thursday
Day[which((t %% 7)    == 3)] <- "F" #friday
Day[which((t %% 7)    == 4)] <- "S" #saturday
Day[which((t %% 7)    == 5)] <- "U" #sunday
Day[which((t %% 7)    == 6)] <- "M" #monday
Day[which((t %% 7)    == 0)] <- "T" #tuesday
Day <- as.factor(Day)

contrasts(Day) # Friday is the default base case

Day <- relevel(as.factor(Day), ref = "U") # Make Sunday the base case

# Build seasonality model with Day as the only predictor
NO2.season <- lm(NO2.ts ~ Day)
summary(NO2.season)

## In order to determine whether there were any seasonal components, we first plotted all the time series data. To check if the seasonality could be captured with trigonometric functions, we constructed a periodogram. However, the periodogram revealed no clear or intuitive periods (the max omega was 192 days). After considering how NO2 emissions primarily come from vehicles, we decided to account for day of the week. All coefficients were statistically significant from the base case (Sunday) except for Saturday, therefore difference between weekend and weekday causes a seasonal variation in the data. Based on this result, we have modeled the seasonal component using Day as a dummy variable.
```

Part 1B - Trends
```{r 1B}

# Build trend model with time as the only predictor
NO2.trend <- lm(NO2.ts ~ t)
summary(NO2.trend) # Time is significant at the 0.001 level

# Plot model with trend
ggplot(dailyAQ.new, aes(x = Group.1,y = NO2.GT.)) + geom_line() +
  stat_smooth(method="lm",col="red") + xlab("") + ylab("Daily Maximum Nitrogen Dioxide Levels")

## There is a trend in the data (increasing), but the trend alone does not appear to be sufficient in explaining N02 levels.

# Diagnostic plots for NO2.trend
autoplot(NO2.trend, labels.id = NULL)
autoplot(NO2.trend, which=4) # No influential points

# There are issues with the Gaussian assumption and non-constant variance.

# Combine trend and season models
NO2.trend.season <- lm(NO2.ts ~ t + Day)
summary(NO2.trend.season) # Coefficient of t is statistically significant

# Plot NO2.trend.season model
ggplot(dailyAQ.new, aes(x = Group.1,y = NO2.GT.)) + geom_line() + 
  geom_line(aes(x=Group.1,y=NO2.trend.season$fitted.values),color="red") +
  xlab("") + ylab("Daily Maximum Nitrogen Levels")

## Accounting for trend and seasonality appears to be a better fit.

# Diagnostic plots for NO2.trend.season
autoplot(NO2.trend.season, labels.id = NULL)
autoplot(NO2.trend.season, which=4) # No influential points

# There are still issues with the Gaussian assumption and non-constant variance

# Compare models based on AIC
AIC(NO2.season)       #4123.301
AIC(NO2.trend)        #4002.334
AIC(NO2.trend.season) #3976.623

# Compare models based on Adjusted R^2
summary(NO2.season)$adj.r.squared #0.04851873
summary(NO2.trend)$adj.r.squared  #0.2966398
summary(NO2.trend.season)$adj.r.squared #0.3522569

## It can be seen that the Adjusted R^2 value significantly increases when trend is added to our NO2 model. NO2.trend.season is the best of these three models based on AIC and adjusted R^2. We will use this model as a starting point moving forward.

```

Part 1C - Autoregressive and Moving Average Components
```{r 1C}

# Regress the NO2.trend.season model over the residuals
e.ts.NO2 <- ts(NO2.trend.season$residuals)

# Plot the residuals over time
autoplot(e.ts.NO2)

# ACF and PACF of the residuals
NO2.acf <- ggAcf(e.ts.NO2)
NO2.pacf <- ggPacf(e.ts.NO2)

# Plot ACF and PACF side by side
ggarrange(NO2.acf,NO2.pacf ,nrow=2,ncol=1)

# The ACF decays sinusoidally
# The PACF cuts off after 2 lags (p = 2)
# These two behavioral characteristics indicate that AR(2) model is appropriate
# Since the ACF doesn't show a slow or linear decary, based on this alone, the time series data is stationary and so there are autoregressive and moving average components.

# Construct an autoregressive model, AR(2)
NO2.ar2 <- arima(e.ts.NO2, order=c(2,0,0))
summary(NO2.ar2)

# Construct a moving average model, MA(1)
NO2.ma1 <- arima(e.ts.NO2, order=c(0,0,1))
summary(NO2.ma1)

# Construct an ARMA model, ARMA(2,1)
NO2.arma21 <- arima(e.ts.NO2, order=c(2,0,1))
summary(NO2.arma21)

# Automatically select models using the arima() function
NO2.auto <- auto.arima(e.ts.NO2, approximation=FALSE)
summary(NO2.auto) # Constructs ARMA(2,1) automatically

## Autoregressive and moving average components were determined by analyzing the ACF and PACF of residuals. Autoregressive and moving average models were then built based on the characteristics depicted in these plots.


```

Part 1D - Model Assessment
```{r 1D}

## Assessment of seasonal model (accounts for day of the week)
summary(NO2.season)$adj.r.squared #0.04851873
AIC(NO2.season) #4123.301

# Construct seasonal model with trigonometric functions for comparison
NO2.trigseason <- lm(NO2.ts ~ sin(2*pi*t*max.omega.NO2) + cos(2*pi*t*max.omega.NO2))
summary(NO2.trigseason)$adj.r.squared #0.07473563
AIC(NO2.trigseason) #4108.625

# NO2.trigseason is slightly better than NO2.season, however both are inadequate

# Season model diagnostics
autoplot(NO2.season) # Non-Gaussian tails
autoplot(NO2.trigseason) # Non-Gaussian tails

library(olsrr)
ols_test_breusch_pagan(NO2.season) # Nonconstant variance (P < 0.05)
ols_test_breusch_pagan(NO2.trigseason) # Nonconstant variance (P < 0.0001)

## Assessment of trend model
summary(NO2.trend)$adj.r.squared #0.2966398
AIC(NO2.trend) #4002.334

# Trend season model from 1B
summary(NO2.trend.season)$adj.r.squared #0.3522569
AIC(NO2.trend.season) #3976.623

anova(NO2.trend, NO2.trend.season) # Partial F Test indicates that season component (Day) adds predictive value (P < 0.0001), thus choose the larger model (NO2.trend.season)

# Construct trend season model with trigonometric functions for comparison
NO2.trend.trigseason <- lm(NO2.ts ~ t + sin(2*pi*t*max.omega.NO2) + cos(2*pi*t*max.omega.NO2))
summary(NO2.trend.trigseason)$adj.r.squared #0.3041614
AIC(NO2.trend.trigseason) #4000.19

anova(NO2.trend, NO2.trend.trigseason) # Partial F Test indicates season component (trig functions) adds some predictive value (P < 0.05)

# NO2.trend.season (Day) is better than NO2.trend.trigseason (trig function) based on adjusted R-squared and AIC. The seasonal component of the Day model is also more intuitive than the trigonometric model, which uses a seemingly arbitrary period (192 days).

# Trend diagnostics
autoplot(NO2.trend.season) # Non-Gaussian tails
ols_test_breusch_pagan(NO2.trend.season) # Nonconstant variance (P = 0.00017)

e.ts.NO2 <- ts(NO2.trend.season$residuals) # Regress NO2.trend.season model over the residuals
Box.test(e.ts.NO2, type="Ljung-Box") # Residual correlations are NOT independent (P < 2.2e-16)

## Assessment of autoregressive and moving average models

# Compare 4 models based on AIC
AIC(NO2.ar2) #AIC=3727.311
AIC(NO2.ma1) #AIC=3814.921
AIC(NO2.arma21) #AIC=3725.118
AIC(NO2.auto) #AIC=3723.118

# Assess residuals vs. fitted
model1 = ggplot() + geom_point(aes(x=fitted(NO2.ar2), y=NO2.ar2$residuals)) + ggtitle("AR2")
model2 = ggplot() + geom_point(aes(x=fitted(NO2.ma1), y=NO2.ma1$residuals)) + ggtitle("MA1")
model3 = ggplot() + geom_point(aes(x=fitted(NO2.arma21), y=NO2.arma21$residuals)) + ggtitle("ARMA21")
model4 = ggplot() + geom_point(aes(x=fitted(NO2.auto), y=NO2.auto$residuals)) + ggtitle("Auto")

ggarrange(model1, model2, model3, model4, ncol=2, nrow=2) # Residuals appear homoscedastic with zero mean

# Assess normality of residuals
model1 = qplot(sample=NO2.ar2$residuals) + stat_qq_line(color="red") + ggtitle("AR2")
model2 = qplot(sample=NO2.ma1$residuals) + stat_qq_line(color="red") + ggtitle("MA1")
model3 = qplot(sample=NO2.arma21$residuals) + stat_qq_line(color="red") + ggtitle("ARMA21")
model4 = qplot(sample=NO2.auto$residuals) + stat_qq_line(color="red") + ggtitle("Auto")

ggarrange(model1, model2, model3, model4, ncol=2, nrow=2) # Residuals show non-Gaussian tails

# Plot diagnostics for independence of residuals using tsdiag()
ggtsdiag(NO2.ar2,gof.lag=20)
Box.test(NO2.ar2$residuals, type="Ljung-Box") # Residuals are independent (P = 0.8175)

ggtsdiag(NO2.ma1,gof.lag=20)
Box.test(NO2.ma1$residuals, type="Ljung-Box") #vResiduals are NOT independent (P < 0.001)

ggtsdiag(NO2.arma21,gof.lag=20)
Box.test(NO2.arma21$residuals, type="Ljung-Box") # Residuals are independent (P = 0.8161)

ggtsdiag(NO2.auto,gof.lag=20)
Box.test(NO2.auto$residuals, type="Ljung-Box") # Residuals are independent (P = 0.8205)

# Plot the autocorrelation (ACF) and partial autocorrelation (PACF) of the residuals of NO2.auto
NO2.auto.resid.acf <- ggAcf(NO2.auto$residuals)
NO2.auto.resid.pacf <- ggPacf(NO2.auto$residuals)
ggarrange(NO2.auto.resid.acf,NO2.auto.resid.pacf,nrow=2,ncol=1) # A few slightly significant lags

## NO2.auto, the selected model, accounts for correlation in the residuals. Aside from the non-Gaussian upper tail, the diagnostics indicate no remaining issues.

```

Part 1E - Forecasting
```{r 1E}

# Forecast the next 7 days of NO2 level residuals
NO2.auto.forecast <- forecast(NO2.auto, h=7)
plot(NO2.auto.forecast)

## Prediction performance

# The test period in days
next.7days.time <- c(length(orig.NO2.ts)-6):(length(orig.NO2.ts))
next.7days.Day <- c("T","W","R","F","S","U","M") #last 7 days start on a Tuesday

# The test data frame
next.7days <- data.frame(t = next.7days.time,
                         Day = next.7days.Day,
                         NO2 = orig.NO2.ts[next.7days.time])

# The actual time series for the test period
next.7days.ts <- ts(next.7days$NO2)

# Prediction for the next 7 days by NO2.auto
E_Y.pred <- predict(NO2.trend.season, newdata=next.7days)
e_t.pred <- forecast(NO2.auto, h=7)
next.7days.prediction <- E_Y.pred + e_t.pred$mean

# MSE
mean((next.7days.prediction - next.7days$NO2)^2) #561.5447

# Plot actual values and predicted values
plot(ts(next.7days$NO2),type='o',ylim=c(0,300))
lines(ts(next.7days.prediction),col='red',type='o')
lines(1:7, E_Y.pred + e_t.pred$lower[,2], col = "red", lty = "dashed")
lines(1:7, E_Y.pred + e_t.pred$upper[,2], col = "red", lty = "dashed")
legend(1,60, legend = c("Actual", "Predicted"), lwd = 2, col = c("black", "red"))

## The MSE value of the 7-day forecast is 561.5447 which is reasonable.

```

Part 2A - Reproduce appearance of time series
```{r 2A}

# Simulate 12 months of daily max NO2 concentration with the chosen model
# Days in 12 months: 7*52 
set.seed(1)
auto.sim <- arima.sim(n=7*52, list(ar=c(NO2.auto$coef[1],NO2.auto$coef[2]),
                                      ma=c(NO2.auto$coef[3])),
                        sd=sqrt(NO2.auto$sigma2))

# Create time variable for next 12 months
next.12mos.time <- c(1:(7*52))

# Create new linear model of max NO2 concentrations vs. the day and the seasonality
NO2.trend.seasonal<-lm(NO2.ts[next.12mos.time]~next.12mos.time+ sin(2*pi*next.12mos.time/7) + cos(2*pi*next.12mos.time/7))
summary(NO2.trend.seasonal)

# Create test data frame
# The test data frame
next.12mos <- data.frame(t = next.12mos.time,Day = next.7days.Day, NO2 = NO2.ts[next.12mos.time])

# The actual time series for the test period
next.12mos.ts <- NO2.ts[next.12mos.time]

next.12mos.ts <- ts(next.12mos$NO2)

# Simulate 12 months
EY.pred <- predict(NO2.trend.seasonal, newdata=next.12mos)
et.pred  <- forecast(NO2.auto, h=364)
next.12mos.prediction <- EY.pred + et.pred$mean

# Plot actual values and predicted values (Observations and Simulation)
plot(ts(next.12mos$NO2),type='o',ylim=c(0,400))
lines(ts(next.12mos.prediction),col='red',type='o')
lines(1:364, EY.pred + et.pred$lower[,2], col = "red", lty = "dashed")
lines(1:364, EY.pred + et.pred$upper[,2], col = "red", lty = "dashed")
legend(1,60, legend = c("Actual", "Predicted"), lwd = 2, col = c("black", "red"))

```

Part 2B - Reproduce observed trends
```{r 2B} 

# Compare linear models of simulations and observations
# Create linear model of trends and seasonality from simulation
Sim.trend.seasonal <- lm(next.12mos.prediction~next.12mos.time+ sin(2*pi*next.12mos.time/7) + cos(2*pi*next.12mos.time/7))

# Compare coefficients of time of Simulation model and Observations model
summary(Sim.trend.seasonal) # 0.2729
summary(NO2.trend.seasonal) # 0.2597

## The percentage difference of the coefficient of time is 4.95% which is quite small.

```

Part 2C - Reproduce seasonality
```{r 2C}

# Comparing periodograms of simulations and observations
# periodogram of simulations
pg.Sim <- spec.pgram(log(next.12mos.prediction), spans = 9, demean=T, log='no')

# periodogram of observations
pg.Obs <- spec.pgram(log(next.12mos.ts), spans = 9, demean=T, log='no')

# The peak in the simulated periodogram is very defined (expected) whereas the original periodogram has no clear period. 

```

Part 2D - Reproduce observed mean and variance
```{r 2D}

# Comparing mean and variance of the simulations and observations
mean(next.12mos.prediction) # 160.1174
mean(next.12mos.ts) # 160.9731
# Percent difference between means is 0.533% (mean of simulation is 0.8557 microg/m^3 smaller than the observations)

var(next.12mos.prediction) # 940.635
var(next.12mos.ts) # 2718.505
# Percent difference between variance is 97.17

# The variance of the simulation is much smaller than the observations

```

Part 2E - Reproduce autocorrelation of time series
```{r 2E}
# Comparing the ACF and PACF of the observations and simulations.
e.ts.sim <- ts(Sim.trend.seasonal$residuals)
e.ts.obs <- ts(NO2.trend.seasonal$residuals)

# ACF plots
ggAcf(e.ts.sim)
ggAcf(e.ts.obs)
# The ACF of the simulation decreases linearly while the ACF of the observation
# decreases sinuisoidally

# PACF plots
ggPacf(e.ts.sim)
ggPacf(e.ts.obs)
# The PACF of the simulation cuts off after one lag quickly and exponentially whereas the PACF for observations shows sinuisoidal behavior and cuts off after lag 15.
```